Inference using context-dependent models of mutation

### 2017-05-12 writeups/context-dependence.tex

\begin{example}[1-D Ising model with Glauber dynamics]
    In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
    imagined as magnetic dipoles,
    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
    Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
    The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
    The following dynamics preserve the stationary distribution:
    each site, independently at rate $\lambda$,
    forgets its spin,
    and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+exp(-(E(x)-E(y)))$,
    this is:
    \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
          $+$  &   $-$   &  $\lambda$ \\
          $-$  &   $+$   &  $\lambda$ 
        \end{tabular} 
        \qquad and \qquad
        \begin{tabular}{cc}
        pattern  &  energy \\
        $+-$ or $-+$  &   $\beta$ \\
        $+$ &   $\gamma$
        \end{tabular} 
    \end{center}
    and 
    \begin{align*}
        \phi(e) = \frac{1}{1+\exp(-e)} .
    \end{align*}

@@

\begin{table}
  \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
          \hline
            x & y & truth & fitted \\ 
          \hline
          \nA  &   \nT           & 0.10 & 0.0996 \\ 
          \nT  &   \nA           & 0.10 & 0.0982 \\ 
          \nC  &   \nG           & 0.15 & 0.1505 \\ 
          \nG  &   \nC           & 0.15 & 0.1497 \\ 
          \nA  &   \nC           & 0.08 & 0.0784 \\ 
          \nT  &   \nG           & 0.08 & 0.0792 \\ 
          \nA  &   \nG           & 0.08 & 0.0798 \\ 
          \nT  &   \nC           & 0.08 & 0.0795 \\ 
          \nC  &   \nA           & 0.12 & 0.1202 \\ 
          \nG  &   \nT           & 0.12 & 0.1196 \\ 
          \nC  &   \nT           & 0.12 & 0.1198 \\ 
          \nG  &   \nA           & 0.12 & 0.1193 \\ 
       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\ 
       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\ 
    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\ 
    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\ 
           \hline
        \end{tabular}
  \end{center}
  \caption{ Results from iterative fitting to a $10^6$ sequence evolved to have 28.5\% sequence divergence
    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$ 
    were constrained to be equal (DNA strand symmetry),
    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
    \label{tab:cpg_results} }
\end{table}

@@

%%%%%%%%%%%
\subsection{TASEP}

Strings of length $10^5$ were evolved under TASEP for varying amounts of time,
and pattern occurrences counted.
Running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
Apparently, the patterns used were not sufficiently independent in this case,
as the posterior is too sharp.

Here's a short example sequence with $\lambda=1$:
\begin{center}
 \setlength{\tabcolsep}{0pt}
\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
X&X&X&O&O&X&O&O&O&O&X&O&X&O&X&X&O&O&X&O&O&X&O&X&X&X&O&X&O&O&O&X&X&O&X&O&O&O&X&X&O&O&O&X&O&O&O&O&X&O&X&X&O&X&X&O&O&X&O&O \\
$\centerdot$&O&X&O&X&O&O&O&X&$\centerdot$&O&X&O&O&X&X&X&o&x&$\centerdot$&o&O&X&x&x&O&X&O&X&o&o&O&X&X&$\centerdot$&$\centerdot$&o&o&O&X&X&o&$\centerdot$&O&X&$\centerdot$&o&o&$\centerdot$&$\centerdot$&O&X&X&x&O&X&o&O&X&$\centerdot$
\end{tabular}
\end{center}
here particles are X's, moving to the right; and spaces are O's;
in the second line,
$\centerdot$ is a position that has not considered a change; lowercase letters are positions that considered a change but did not;
and upper case are positions that did change.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{writeup-plots/tasep_all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates for six different data sets of the parameter in TASEP (mutation rate multiplied by time).
  Vertical line shows the true value.
  }
\end{figure}

\subsection{Ising model}

Strings of length $10^6$ were evolved under Glauber dynamics on the Ising model with varying parameters,
and pattern occurrences counted.
As for TASEP, running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
In this case, the true parameter values have good distribution within the estimated posterior distributions.

Here is an example, with all parameters equal to 1, for $t=.5$:
\begin{center}
 \setlength{\tabcolsep}{0pt}
\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
X&X&O&X&X&O&O&X&O&X&O&X&X&O&O&X&O&O&O&O&X&X&X&O&X&O&X&O&X&O&O&O&X&X&O&X&X&X&X&X&O&X&O&X&O&X&X&X&X&O&O&X&X&X&X&X&X&O&O&O \\
x&x&o&x&X&X&O&x&o&x&$\centerdot$&x&x&O&X&X&$\centerdot$&O&X&O&X&x&x&o&x&o&x&o&x&O&X&O&$\centerdot$&$\centerdot$&O&O&X&X&O&X&X&X&o&x&o&$\centerdot$&X&O&X&o&o&x&X&X&X&O&X&o&$\centerdot$&$\centerdot$
\end{tabular}
\end{center}
The notation is as above; except that the entire matched pattern involved in a change is uppercase:
to simulate, we look in windows of size 3; so if XOO changes to XXO, the whole thing is uppercase, not just the middle X.

\begin{figure}
  \begin{center}
    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
  \end{center}
  \caption{
  MCMC traces of the parameters in the Ising model across 100,000 iterations.
  The units of $\lambda$ are in mean number of changes per site.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=1]{writeup-plots/ising_all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the time-scaled mutation rate parameter $\lambda t$,
  in ten different simulated datasets under the Ising model.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=2]{writeup-plots/ising_all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the inverse temperature $\beta$ in the Ising model.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=3]{writeup-plots/ising_all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the magnetization parameter $\beta$ in the Ising model.
  Vertical line shows the true value.
  }
\end{figure}

%%%
\subsection{CpG mutation, on a tree}

Nucleotide sequences of length $10^6$ were simulated down two (unequal) branches of a tree,
with all single-base substitution rates $m_{xy}$ set equal (varying from a total of .01 to .1 across the tree)
and an additional CpG rate $\gamma$ set to either 0 or three times the single-base rate.
Analysis was done with $\ell=2$ and $w=1$ (a five-to-one window).

The state at the root of the tree was independently chosen bases with equal frequencies,
and these frequencies at the root were added as parameters to the model (with a Dirichlet prior).

Here is a short example, with all parameters equal to 1, with each branch of the tree having length $0.1$:
\begin{center}
 \setlength{\tabcolsep}{0pt}
\begin{tabular}{cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc}
T&G&G&A&G&A&G&G&T&C&T&G&C&C&A&T&A&G&A&C&G&G&G&C&G&C&C&G&C&C&T&A&G&G&A&G&C&C&G&C&G&T&T&T&T&G&C&G&T&G&G&T&G&T&T&G&T&G&G&C \\
$\centerdot$&$\centerdot$&g&C&G&a&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&A&C&A&$\centerdot$&$\centerdot$&$\centerdot$&T&A&T&$\centerdot$&$\centerdot$&$\centerdot$&T&G&c&C&C&A&C&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&G&G&t&A&T&$\centerdot$&$\centerdot$&G&A&g&C&G&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&t&T&G&A \\
$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&T&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&T&T&G&T&$\centerdot$&$\centerdot$&G&A&T&$\centerdot$&$\centerdot$&$\centerdot$&G&G&c&t&A&A&G&C&$\centerdot$&$\centerdot$&$\centerdot$&G&T&G&A&t&t&t&g&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&G&A&$\centerdot$&$\centerdot$&$\centerdot$&$\centerdot$&A&C
\end{tabular}
\end{center}
Top is ancestral sequence (``unobserved''); bottom two are at the tips.

The point estimates are good,
as are the MCMC estimates,
but there seems to be mixing trouble having to do with the root frequencies.

\begin{figure}
  \begin{center}
    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
  \end{center}
  \caption{
  MCMC traces across 660,000 iterations of the frequencies at the root of the (two-taxon) tree in the CpG model.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
  \end{center}
  \caption{
  MCMC traces across 660,000 iterations of several mutation rate parameters in the CpG model.
  Units are in mean number of substitutions per base across the entire tree.
  Vertical line shows the true value.
  }
\end{figure}

%%%%%%%%
\section{Power sims}

Assess power and accuracy in the Ising model, by running replicates at these parameters:

\begin{tabular}{|rlllrrr|}
  \hline
  seqlen   &  $\lambda t$   &  $\beta$  &  $\gamma$  &  $w$ &  $\ell$   & \# reps \\
  \hline
  $10^4$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
  $10^3$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
  $10^2$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
  \hline
  $10^4$    &   .4  &   1   &   .5  &   3   &   3   &   10  \\
  $10^4$    &   .1  &   1   &   .5  &   3   &   3   &   10  \\
  $10^4$    &   .05  &   1   &   .5  &   3   &   3   &   10  \\
  \hline
  $10^4$    &   .2  &   1   &   0  &   3   &   3   &   10  \\
  \hline
  $10^4$    &   .2  &   1   &   .5  &   3   &   2   &   10  \\
  $10^4$    &   .2  &   1   &   .5  &   3   &   1   &   10  \\
  $10^4$    &   .2  &   1   &   .5  &   3   &   0   &   10  \\
  \hline
\end{tabular}

@@

\begin{itemize}

  \item estimating freq spectrum [Hernandez]

  \item probabilistic ancestral states for common ancestor of e.g. humans 4N generations ago [Siepel] -- would widely used for primates

  \item ancestral CpG

  \item ancient DNA

  \item noisy image transmission denoising: first learn model; then reduce redundancy

  \item root distribution and confounding/mixing

  \item shrinkage prior on highly parameterized model

  \item Show that taking no context ($\ell=0$) can be positively misleading in some context.

  \item How to identify motifs in the residuals?

\end{itemize}

\subsection{Acknowledgements}
Erick, Jessica, Graham, Yaniv, Sergey, Simon, Matt, Rasmus

>>>>

\begin{example}[Gibbs sampling of the Ising model]
    In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
    imagined as magnetic dipoles,
    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
    Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
    The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
    The following process, also known as ``Glauber dynamics'', preserves the stationary distribution:
    each site, independently at rate $\lambda$,
    forgets its spin,
    and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+exp(-(E(x)-E(y)))$,
    this is:
    \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
          $+$  &   $-$   &  $\lambda$ \\
          $-$  &   $+$   &  $\lambda$ 
        \end{tabular} 
        \qquad and \qquad
        \begin{tabular}{cc}
        pattern  &  energy \\
        $+-$ or $-+$  &   $\beta$ \\
        $+$ &   $\gamma$
        \end{tabular} 
    \end{center}
    and 
    \begin{align*}
        \phi(e) = \frac{1}{1+\exp(-e)} .
    \end{align*}

@@

\begin{table}
  \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
          \hline
            x & y & truth & MLE \\ 
          \hline
          \nA  &   \nT           & 0.10 & 0.0996 \\ 
          \nT  &   \nA           & 0.10 & 0.0982 \\ 
          \nC  &   \nG           & 0.15 & 0.1505 \\ 
          \nG  &   \nC           & 0.15 & 0.1497 \\ 
          \nA  &   \nC           & 0.08 & 0.0784 \\ 
          \nT  &   \nG           & 0.08 & 0.0792 \\ 
          \nA  &   \nG           & 0.08 & 0.0798 \\ 
          \nT  &   \nC           & 0.08 & 0.0795 \\ 
          \nC  &   \nA           & 0.12 & 0.1202 \\ 
          \nG  &   \nT           & 0.12 & 0.1196 \\ 
          \nC  &   \nT           & 0.12 & 0.1198 \\ 
          \nG  &   \nA           & 0.12 & 0.1193 \\ 
       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\ 
       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\ 
    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\ 
    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\ 
           \hline
        \end{tabular}
  \end{center}
  \caption{ Results from iterative fitting to a $10^6$ sequence evolved to have 28.5\% sequence divergence
    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$ 
    were constrained to be equal (DNA strand symmetry),
    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
    \label{tab:cpg_results} }
\end{table}

@@

@@

This paper presents a tractable, provably correct method
for doing statistical inference using observations of a particle model
in which probabilities of change are determined only by nearby sites.
Such models appear in many fields, for instance,
as the dynamic counterparts to Markov random fields;
many spin glass models;
and models of DNA sequence evolution.

The methods described here provide a way to efficiently calculate full likelihoods,
which in many circumstances provide the optimal basis for inference \citep{neymanpearson},
fast enough to allow Bayesian inference from Monte Carlo sampling.
The method is fast thanks to its dependence only only local neighborhoods
(so the time does not scale with sequence length once T-mer abundances are counted)
and its reliance on sparse matrices whose structure can be precomputed and cached.
It avoids approximations made by previous methods thanks to the use of T-mers
and explict quantification of the necessary length of the ``arms'' of the T-mers.
We first discuss possible uses of the method,
followed by challenges (and possible solutions).

\paragraph{Other applications}
One class of applications stems from \emph{inference of process} --
where the numerical values of various mutation rates are directly of interest.
For instance, CpG hypermutability depends on the \nC\nG pair being methylated,
an epigenetic modification that has important consequences for gene regulation;
therefore, inference of CpG mutation rates in ancient taxa provides an indirect estimate
of methylation status in extinct organisms.
Similarly, many mutational motifs are thought to be caused by particular enzymes;
inference of relative rates of these motifs
may provide clues as to the activity and evolution of those enzymes.
Furthermore, iterative model building by examining residual motif counts
could provide a powerful way to discover unknown processes
whose action had been obscured by other sources of noise.

In other applications, reconstruction of an unknown sequence may of interest.
For instance, training data might be used to estimate parameters of a noisy transmission line,
facilitating the later reconstruction of noisily transmitted signals.
Alternatively, multiple observations of independently transmitted copies of a single sequence
(e.g., down a phylogenetic tree)
might be used to estimate the process, and from this infer the original (ancestral) sequence
(or sample from its posterior distribution).

\paragraph{Limitations}
The model presented assumes that the same process occurs at every location along the sequence,
and that the length of the sequnce does not change (no indels).
Neither of these assumptions are likely to be true for any data set of DNA sequence;
however, both assumptions are also made the state of the art in phylogenetics.
Some relaxation of the assumptions is possible -- like in phylogenetics \citep{ratemixtures},
we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor.
Short deletions might be accomodated by introducing a ``gap'' letter to the alphabet,
but insertions may be problematic.

\paragraph{Computation}
The main factor that determines computation time
is the lengths of the T-mers used for inference,
since all computations scale with the number of possible patterns at the long end of the T-mer.
Although the model is efficent enough to deal with fairly long T-mers,
this exponential growth puts hard constraints on the application of the method.
The length of T-mers required to accurately compute the likelihood in most models grows at most linearly
with the mean density of substitution between the initial and final sequence.
This does not in practice greatly increase the length of required T-mer 
since substitution densities approaching 100\% are unlikely to be usable in practice
especially if how the sequences align with each other is unknown.
A more serious determining factor of T-mer length is the degree of dependency in the model itself.
For instance, inserting the 13-base pair motif selected against by meiotic drive in primates \citep{myers2010drive}
would require T-mers of width at least 25, and hence vectors of length $4^25 \approx 1.13 \times 10^15$.

\paragraph{Modeling}
Careful building of complex models in this framework may take care,
but no more than in any other class of models (e.g., multivariate linear regression).
It is easy to specify nonidentifiable models -- 
for instance, all 2mer mutation motifs include the 1mer motifs.
The large number of possible mutation motifs also makes multiple testing an issue;
an unguided approach might include all possible motifs but place a shrinkage prior on their values \citep{shrinkage,horseshoe}.
The converse approach of adding motifs that are overrepresented in the residuals
may be quicker, but entails some degree of arbitrariness in choosing the motifs to add (and in which order).
Finally, inference on trees may be difficult unless the root placement and distribution are strongly constrained:
otherwise, initial investigations suggest the likelihood surface is often strongly ridged --
for instance, increasing the abundance of \nG nucleotides at the root while also increasing the mutation rate away from \nG
can achieve roughly the same result.
However, in phylogenetic applications outgroups can be used to constrain these quantities.

\subsection{Acknowledgements}
The initial impetus for this project came from Matt Dean; 
along the way substantial encouragement and useful suggestions were provided by
Erick Matsen, Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
In particular, Erick Matsen helped improve the code and suggested the name ``T-mer''.
Many thanks are also due Jessica Crisci for curating the human-chimp data.
